An elaborate data set on human gait and the effect of mechanical perturbations

### 2015-03-16 paper.tex
Address's most of Ton's final comments. Other's will be addressed in subsequent commits/PRs.

There are some notable gait data sets and databases besides Winter's
authoritative set that are publicly available. The International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and mostly unencumbered
use. For example, the kinematic and force plate measurements from several
subjects from \cite{Vaughan1992} is available on the site. At another website,
the CGA Normative Gait Database, \cite{Kirtley2014} curates and shares
normative clinical gait data collected from multiple labs and these datasets
have influenced other studies, for example the average gait cycles from
children used in \cite{Bogert2003}.

@@

Our use case for the data is centered around the need for bio-inspired control
systems in emerging powered prosthetics and orthotics. Ideally, a powered
prosthetic would behave in such a way that the user would feel like their limb
was never disabled. There are a variety of approaches to developing
bio-inspired control systems, some of which aim to mimic the reactions and
motion of an able-bodied person. A modern gait lab is able to collect a variety
of kinematic, kinetic, and physiological data from humans during gait. This
data can potentially be used to drive the design of the human-mimicking
controller. With a rich enough data set, one may be able to identify control
mechanisms used during a human's natural gait and recovery from perturbations.
We hypothesize that by forcing the human to recover from external
perturbations, the resulting reactive actions can be used along with system
identification techniques to expose the feedback related relationships among
the human's sensors and actuators. We have collected data that is richer than
previous gait data sets and may be rich enough for control identification. The
data can also be used for verification purposes for controllers that have been
designed in other manners, such as those constructed from first principles
\citep{Geyer2010}.

With all of this in mind, we collected over seven and half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The protocol for the majority of the trials included two minutes
of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing \citep{White2013}.

Furthermore, we include a small Apache licensed open source software library
for basic gait analysis and demonstrate its use in the paper. The combination
of the open data and open software allow the results presented within to be
computationally reproducible and instructions are included in the associated
repository for doing so.

@@

\subsection*{Equipment}
%
The data were collected in the Laboratory for Human Motion and Control at
Cleveland State University, using the following equipment:
%
\begin{itemize}
  \item A R-Mill treadmill which has dual 6 degree of freedom force plates,
    independent belts for each foot, and lateral/pitch motion capabilities
    (Forcelink, Culemborg, Netherlands).
  \item A 10 Osprey camera motion capture system paired with the Cortex
    3.1.1.1290 software (Motion Analysis, Santa Rosa, CA, USA).
  \item USB-6255 data acquisition unit (National Instruments, Austin, Texas,
    USA).
  \item Four ADXL330 Triple Axis Accelerometer Breakout boards attached to the
    treadmill (Sparkfun, Niwot, Colorado, USA).
  \item D-Flow software (versions 3.16.1 to 3.16.2) and visual display system,
    (Motek Medical, Amsterdam, Netherlands).
\end{itemize}

The Cortex software delivers high accuracy 3D marker trajectories from the
cameras along with data from force plates and analog sensors
(EMG/Accelerometer) through a National Instruments USB-6255 data acquisition
unit. D-Flow is required to collect data from any digital sensors and to
control the treadmill's motion (lateral, pitch, and belts). D-Flow can process
the data in real time and/or export data to file.

@@

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that is detectable in the force measurements,
likely due to the flexible undercarriage and pitch motion mechanism. Trials 6-8
are affected by this vibration mode. During trials 9-15 the treadmill was
stabilized with wooden blocks. During the remaining trials the treadmill was
stabilized with metal supports. See the Data Limitations Section for more
details.

@@

\subsection*{Protocol}
%
The experimental protocol consisted of both static measurements and walking on
the treadmill for 10 minutes under unperturbed and perturbed conditions. Before
a set of trials on the same day, the motion capture system was calibrated using
the manufacturer's recommended procedure. Before each subject's gait data were
collected, the subject changed into athletic shoes, shorts, a sports bra,
a baseball cap~\footnote{A cap was used to avoid having to shave participants'
hair but at the expense of accuracy.}, and a rock climbing harness. All 47 markers
were applied directly to the skin at the landmarks noted in
Table~\ref{tab:marker-labels} except for the heel, toe, and head markers, which
were placed on the respective article of clothing.~\footnote{The sacrum and
rear pelvic markers may have been placed on the shorts for a small number of
the subjects.} Then the subject self-reported their age, gender, and mass.
Finally, their height was measured by the experimentalist and four reference
photographs (front, back, right, left) were taken of subject's marker
locations.

After obtaining informed consent and a briefing by the experimentalist on the
trial protocol, the subject followed the verbal instructions of the
experimentalist and the on-screen instructions from the video display. The
protocol for a single trial was as follows:
%
\begin{enumerate}
  \item Subject stepped onto the treadmill and markers were identified with
    Cortex.
  \item The safety rope was attached loosely to the rock climbing harness such
    that no undue forces were acting on the subject during walking, but that
    the harness would prevent a full fall.
  \item The subject started by stepping on sides of treadmill so that feet did
    not touch the force plates and the force plate signals are zeroed. This
    corresponds to the ``Force Plate Zeroing'' event.
  \item Once notified by the video display, the subject stood in the
    calibration pose: standing straight up, looking forward, arms out by their
    sides (approximately 45 degree abduction) and the event, ``Calibration
    Pose'', was manually recorded by the operator.
  \item A countdown to the first normal walking phase was displayed. At the end
    of the countdown the event ``First Normal Walking'' was recorded and the
    treadmill ramped up to the specified speed and the subject was instructed
    to walk normally, to focus on the ``endless'' road on the display, and not
    to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begun and was recorded as ``Longitudinal Perturbation''.
  \item After 8 minutes of walking under the influence of the perturbations,
    the second normal walking phase begun and was recorded as ``Second Normal
    Walking''.
  \item After 1 minute of normal walking, a countdown was shown on the display
    and the treadmill decelerated to a stop.
  \item The subject was instructed to step off of the force plates for 10
    seconds and the ``Unloaded End'' event was recorded.
  \item The subject could then take a rest break before each additional trial.
\end{enumerate}

To create the signals, we started by generating random 100~\si{\hertz}
acceleration signals using the Simulink discrete-time Gaussian white noise
block followed by a saturation block set at the maximum belt acceleration of
15~\si{\meter\per\second\squared}. The signal was then integrated to obtain
belt speed and high-pass filtered with a second-order Butterworth filter to
eliminate drift. One of the three mean speeds were then added to the signal and
limited between \SIrange{0}{3.6}{\meter\per\second}. The cutoff frequencies of
the high-pass filter, as well as the variance in the acceleration signal, were
manually adjusted until acceptable standard deviations for each mean speed were
obtained: \SIlist{0.06;0.12;0.21}{\meter\per\second} for the three speeds,
respectively. These ensured that the test subjects were sufficiently perturbed
at each speed, while remaining within the limits of our equipment and testing
protocol.

To ensure that the treadmill belts could accelerate to the desired values, the
high performance mode in the D-Flow software was enabled. This had the side
effect of enabling too rapid of accelerations when the belt speed changed to or
from zero speed. To eliminate this, a suitable ramped acceleration and
deceleration were generated for the speed transitions.

The MATLAB script and Simulink model produce a comma-delimited text file of
five desired belt speed signals indexed by the time stamp. There are slow, normal,
and fast walking speeds and slow and fast running speeds.~\footnote{The running
signals were not used in the experiments presented in this paper.}

Figure~\ref{fig:input_output} gives an idea of the effect of the treadmill and
controller dynamics by plotting the measured speed of the treadmill belts from
loaded trials (76, 77, 78) against the commanded treadmill control input
signal. The system introduces a delay and seems to act as a low pass filter.
The standard deviations of the measured speeds do not significantly differ from
the desired values: \SIlist{0.05;0.12;0.2}{\meter\per\second} for the three
speeds, respectively.
%
\begin{figure}
  \centering
  \includegraphics{figures/input_vs_output.pdf}
  \cprotect\caption{Commanded treadmill belt speed (red) and the recorded speed
    (blue) for average belt speeds of \SIlist{0.8;1.2;1.6}{\meter\per\second},
    respectively. Generated by \verb|src/input_output_plot.m|.}
  \label{fig:input_output}
\end{figure}

We make use of the full body 47 marker set described in \cite{Bogert2013} and
presented in detail in Table \ref{tab:marker-labels}. As with all camera based
motion capture systems, the markers sometimes go missing in the recording. When
a marker goes missing, if the data was recorded in a D-Flow version less than
3.16.2rc4, D-Flow continues to record the last non-missing value in all
three axes until the marker is visible again. In D-Flow versions greater than
or equal to 3.16.2rc4, the missing markers are indicated in the TSV file as
either \verb|0.000000| or \verb|-0.000000|. The D-Flow version must be provided
in the meta data YAML file to be able to distinguish this detail.

@@

>>>>

There are some notable gait data sets and databases besides Winter's
authoritative set that are publicly available. The International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and mostly unencumbered
use. For example, the kinematic and force plate measurements from several
subjects from \cite{Vaughan1992} is available on the site. At another website,
the CGA Normative Gait Database, \cite{Kirtley2014} curates and shares
normative clinical gait data collected from multiple labs and these datasets
have influenced other studies, for example \cite{Bogert2003} made use of the
average gait cycles from the child subjects.

@@

Our use case for the data is centered around the need for bio-inspired control
systems in emerging powered prosthetics and orthotics. Ideally, a powered
prosthetic would behave in such a way that the user would feel like their limb
was never disabled. There are a variety of approaches to developing
bio-inspired control systems, some of which aim to mimic the reactions and
motion of an able-bodied person. A modern gait lab is able to collect a variety
of kinematic, kinetic, and physiological data from humans during gait. This
data can potentially be used to drive the design of the human-mimicking
controller. With a rich enough data set, one may be able to identify control
mechanisms used during a human's natural gait and recovery from perturbations.
We hypothesize that by forcing the human to recover from external
perturbations, the resulting reactive actions can be used along with system
identification techniques to expose the feedback related relationships among
the human's sensors and actuators. With this in mind, we have collected data
that is richer than previous gait data sets and may be rich enough for control
identification. The data can also be used for verification purposes for
controllers that have been designed in other manners, such as those constructed
from first principles \citep{Geyer2010}.

With all of this in mind, we collected over seven and a half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The final protocol for the majority of the trials included two
minutes of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing
\citep{White2013}.

Furthermore, we include a small Apache licensed open source software library
for basic gait analysis and demonstrate its use in the paper. The combination
of the open data and open software allow the results presented within to be
computationally reproducible and instructions are included in the associated
repository~\footnote{https://github.com/csu-hmc/perturbed-data-paper} for
reproducing the results.

@@

\subsection*{Equipment}
%
The data were collected in the Laboratory for Human Motion and Control at
Cleveland State University, using the following equipment:
%
\begin{itemize}
  \item A R-Mill treadmill which has dual 6 degree of freedom force plates,
    independent belts for each foot, along with lateral translation and pitch
    rotation capabilities (Forcelink, Culemborg, Netherlands).
  \item A 10 Osprey camera motion capture system paired with the Cortex
    3.1.1.1290 software (Motion Analysis, Santa Rosa, CA, USA).
  \item USB-6255 data acquisition unit (National Instruments, Austin, Texas,
    USA).
  \item Four ADXL330 Triple Axis Accelerometer Breakout boards attached to the
    treadmill (Sparkfun, Niwot, Colorado, USA).
  \item D-Flow software (versions 3.16.1 to 3.16.2) and visual display system,
    (Motek Medical, Amsterdam, Netherlands).
\end{itemize}

The Cortex software delivers high accuracy 3D marker trajectories from the
cameras along with data from the force plates and analog sensors (e.g.
EMG/Accelerometer) through a National Instruments USB-6255 data acquisition
unit. D-Flow then receives streaming data from Cortex and any other digital
sensors. It is also responsible for controlling the treadmill's motion
(lateral, pitch, belts). D-Flow can process the data in real time and/or export
data to file.

@@

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that was detectable in the force
measurements, likely due to the flexible undercarriage and pitch motion
mechanism. Trials 6-8 are affected by this vibration mode. During trials 9-15
the treadmill was stabilized with wooden blocks. During the remaining trials
the treadmill was stabilized with metal supports; both in-house fabricated ones
and vendor supplied ones. These supports aimed to improve the stiffness and
frequency response of the force plate system. See the Data Limitations Section
for more details.

@@

\subsection*{Protocol}
The experimental protocol consisted of both static measurements and walking on
the treadmill for 10 minutes under unperturbed and perturbed conditions. Before
a set of trials on the same day, the motion capture system was calibrated using
the manufacturer's recommended procedure. Before each subject's gait data were
collected, the subject changed into athletic shoes, shorts, a sports bra, a
baseball cap~\footnote{A cap was used to avoid having to shave participants'
hair but at the expense of accuracy.}, and a rock climbing harness. All 47
markers were applied directly to the skin at the landmarks noted in
Table~\ref{tab:marker-labels} except for the heel, toe, and head markers, which
were placed on the respective article of clothing.~\footnote{The sacrum and
rear pelvic markers were placed on the shorts for a small number of the
subjects.} Then the subject self-reported their age, gender, and mass.
Finally, their height was measured by the experimentalist and four reference
photographs (front, back, right, left) were taken of subject's marker
locations.

@@

After obtaining informed consent and a briefing by the experimentalist on the
trial protocol, the subject followed the verbal instructions of the
experimentalist and the on-screen instructions from the video display. The
protocol for a single trial was as follows:
%
\begin{enumerate}
  \item Subject stepped onto the treadmill and markers were identified with
    Cortex.
  \item The safety rope was attached loosely to the rock climbing harness such
    that no forces were acting on the subject during walking, but that the
    harness would prevent a full fall.
  \item The subject started by stepping on sides of treadmill so that feet did
    not touch the force plates and the force plate signals are zeroed. This
    corresponds to the ``Force Plate Zeroing'' event.
  \item Once notified by the video display, the subject stood in the
    calibration pose: standing straight up, looking forward, arms out by their
    sides (approximately 45 degree abduction) and the event, ``Calibration
    Pose'', was manually recorded by the operator.
  \item A countdown to the first normal walking phase was displayed. At the end
    of the countdown the event ``First Normal Walking'' was recorded and the
    treadmill ramped up to the specified speed and the subject was instructed
    to walk normally, to focus on the ``endless'' road on the display, and not
    to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begun and was recorded as ``Longitudinal Perturbation''.
  \item After 8 minutes of walking under the influence of the perturbations,
    the second normal walking phase begun and was recorded as ``Second Normal
    Walking''.
  \item After 1 minute of normal walking, a countdown was shown on the display
    and the treadmill decelerated to a stop.
  \item The subject was instructed to step off of the force plates for 10
    seconds and the ``Unloaded End'' event was recorded.
  \item The subject could then take a rest break before each additional trial.
\end{enumerate}

To create the signals, we started by generating random acceleration signals,
sampled at 100~\si{\hertz}, using the Simulink discrete-time Gaussian white
noise block followed by a saturation block set at the maximum belt acceleration
of 15~\si{\meter\per\second\squared}. The signal was then integrated to obtain
belt speed and high-pass filtered with a second-order Butterworth filter to
eliminate drift. One of the three mean speeds were then added to the signal and
limited between \SIrange{0}{3.6}{\meter\per\second}. The cutoff frequencies of
the high-pass filter, as well as the variance in the acceleration signal, were
manually adjusted until acceptable standard deviations for each mean speed were
obtained: \SIlist{0.06;0.12;0.21}{\meter\per\second} for the three speeds,
respectively. These ensured that the test subjects were sufficiently perturbed
at each speed, while remaining within the limits of our equipment and testing
protocol. To ensure that the treadmill belts could accelerate to the desired
values, the high performance mode in the D-Flow software was enabled. The
MATLAB script and Simulink model produce a comma-delimited text file of with
the desired belt speed signals indexed by the time stamp.

Figure~\ref{fig:input_output} gives an idea of the effect of the treadmill and
controller dynamics by plotting the measured speed of the treadmill belts from
loaded trials (76, 77, 78) against the commanded treadmill control input
signal. The system introduces a delay and seems to act as a low pass filter.
The standard deviations of the measured speeds do not significantly differ from
those of the commanded speeds: \SIlist{0.05;0.12;0.2}{\meter\per\second} for
the three speeds, respectively.
%
\begin{figure}
  \centering
  \includegraphics{figures/input_vs_output.pdf}
  \cprotect\caption{Commanded treadmill belt speed (red) and the recorded speed
    (blue) for average belt speeds of \SIlist{0.8;1.2;1.6}{\meter\per\second},
    respectively. Generated by \verb|src/input_output_plot.m|.}
  \label{fig:input_output}
\end{figure}

We make use of the full body 47 marker set described in \cite{Bogert2013} and
presented in detail in Table \ref{tab:marker-labels}. As with all camera based
motion capture systems, the markers sometimes go missing in the recording. When
a marker goes missing, if the data was recorded in a D-Flow version less than
3.16.2rc4, D-Flow continues to record the last non-missing value in all
three axes until the marker is visible again. In D-Flow versions greater than
or equal to 3.16.2rc4, the missing markers are indicated in the TSV file as
either \verb|0.000000| or \verb|-0.000000|. The D-Flow version must be provided
in the meta data YAML file to be able to distinguish this detail.
